---
title: "Gov 50 Final Project"
author: "Jose Valdez Genao"
description: "Factors Influencing School Performance in the Dominican Republic"
output:
  distill::distill_article:
    self_contained: false
---

# Exploring the Correlation Between School Performance in Dominican National Assessments and School Location and Type

## Introduction

This project explores  the performance of schools in the Dominican Republic's National Assessments (Pruebas Nacionales). Every year the Ministry of Education conducts these examinations nationwide, which every student is required to take to graduate from High School. These results are useful to better understand how schools are performing in the different subjects (Spanish, Math, Social Sciences, Natural Sciences).

I aim to understand whether school performance relates to students' location and schooling type. My hypothesis is that urban schools score higher than rural counterparts and that different schooling methods might yield varying results. The rationale behind this hypothesis stems from observations of the Dominican education system and its challenges. Firstly, the hypothesis posits that urban students may outperform their rural counterparts due to differing access to educational resources. Urban areas can enjoy better infrastructure, including more access to electricity, internet, or technology. These advantages can significantly impact students' learning experiences and potentially lead to higher academic performance. Secondly, the hypothesis anticipates performance variations across different school types. Variations are always expected, but in this case, this expectation is rooted in the observation that specific school types, such as academic, or arts-focused schools, might encounter distinct challenges and resources. For instance, technical schools might have access to better-trained teachers, whereas arts schools might face limitations in these aspects due to being more experimental. This analysis is essential given the Dominican Republic's education challenges. Internationally, the country ranked in the lowest performance out of 79 participant countries/economies in the international assessment PISA 2018. Nationally, resource disparities and teacher shortages affect rural areas disproportionately. Conversely, educators in urban areas can benefit from better training due to accessibility to better-funded universities. Addressing these disparities can help understand educational disparities within the country. 


## Data Section

```{r, echo = FALSE}
library(readr)
DR_education <- read_csv('Results_PN_Data.csv')
```

The dataset utilized for this project holds national examination results spanning the years 2016 to 2020, collected within the Dominican Republic. The unit of analysis focuses on school-level mean performance, resulting in a dataset with 34,405 observations. The primary data source consists of these national examinations, which serve as the key dependent variables.

To refine the analysis, one of the independent variable "Regional" was modified to compare school differences within six  selected regions, three urban and three rural provinces. This adjustment was needed due to the absence of a comprehensive national-level map distinguishing urban and rural areas. The selected urban regions for comparison include the National District, Santo Domingo, and Santiago, representing three of the country's most modern and urbanized provinces. Conversely, the chosen rural regions encompass Barahona, San Juan, and Mao, provinces characterized by lower population density and predominantly rural settings. The second independent variable, "Nivel/Modalidad" (level/modality), categorizes each school nationwide into distinct types of schooling, including General, Academic, Professional Technical, or Arts (e.g., School X = Arts).

The outcome variables of interest are the mean scores in four academic subjects: Spanish, Math, Social Sciences, and Natural Sciences. Each academic subject represents a separate outcome variable, resulting in a total of four outcome variables. These variables measure scores on a 30% scale and summarize the mean performance of all students within each school. For example, the mean scores of all students in School X are coded for Spanish (e.g., Spanish = 16.53).

The research design for this study is cross-sectional, as it analyzes the data collected from 2016 to 2020 as a single dataset to explore correlations between the aforementioned variables. 

```{r, echo = FALSE}

library(ggplot2)
library(tidyr)

boxplot_data <- HS_DR_Education %>%
  select(Español, Matemáticas, Sociales, Naturales)

boxplot_data_long <- boxplot_data %>%
  pivot_longer(cols = everything(), names_to = "Subject", values_to = "Mean_Scores")

custom_labels <- c(
  Español = "Spanish",
  Matemáticas = "Math",
  Sociales = "Social Sciences",
  Naturales = "Natural Sciences")

ggplot(boxplot_data_long, aes(x = Subject, y = Mean_Scores)) +
  geom_boxplot() +
  labs(
    x = "Academic Subject",
    y = "Mean Scores",
    title = "National Distribution of Mean Scores in Academic Subjects"
  ) +
  scale_x_discrete(labels = custom_labels)

```


## Results Section
### National Mean Results per Schooling Type

```{r, echo = FALSE}
library(dplyr)
library(knitr)

HS_DR_Education <- DR_education %>%
  filter(!grepl("basica", `Nivel/Modalidad`, ignore.case = TRUE))

HS_DR_Education <- HS_DR_Education %>%
  mutate(Español = as.numeric(gsub("[^0-9.]", "", Español)),
         Matemáticas = as.numeric(gsub("[^0-9.]", "", Matemáticas)),
         Sociales = as.numeric(gsub("[^0-9.]", "", Sociales)),
         Naturales = as.numeric(gsub("[^0-9.]", "", Naturales)))

HS_DR_Education[is.na(HS_DR_Education)] <- 0

results_means <- HS_DR_Education %>%
  group_by(`Nivel/Modalidad`) %>%
  summarize(Spanish = mean(Español, na.rm = TRUE),
            Math = mean(Matemáticas, na.rm = TRUE),
            Social_Sciences = mean(Sociales, na.rm = TRUE),
            Natural_Sciences = mean(Naturales, na.rm = TRUE)) %>%
  filter(`Nivel/Modalidad` %in% c("ADULTOS FORMAL", "ARTES", "MODALIDAD GENERAL", "TECNICO PROFESIONAL", "ACADEMICOS")) %>%
  mutate(Schooling_Type = case_when(
    `Nivel/Modalidad` == "ADULTOS FORMAL" ~ "Adults School",
    `Nivel/Modalidad` == "ARTES" ~ "Arts",
    `Nivel/Modalidad` == "MODALIDAD GENERAL" ~ "General Education",
    `Nivel/Modalidad` == "TECNICO PROFESIONAL" ~ "Technical-Professional",
    `Nivel/Modalidad` == "ACADEMICOS" ~ "Academic"
  )) %>%
  select(Schooling_Type, Spanish, Math, Social_Sciences, Natural_Sciences)

colnames(results_means) <- c("Schooling Type", "Spanish", "Math", "Social Sciences", "Natural Sciences")

kable(results_means, caption = "National Mean Results per Schooling Type")

```


```{r, echo = FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
library(tidytext)

data <- tribble(
  ~Schooling_type, ~Spanish, ~Mathematics, ~Social_Sciences, ~Natural_Sciences,
  "ACADEMIC", 19.902198, 16.816484, 18.340049, 17.725397,
  "ADULTS SCHOOL", 9.828218, 9.661577, 9.290876, 9.042828,
  "ARTS", 16.453333, 14.667868, 15.084307, 14.985406,
  "GENERAL EDUCATION", 15.855032, 15.471510, 15.611412, 15.590785,
  "TECHNICAL-PROFESSIONAL", 18.298454, 17.115917, 16.889827, 17.007560
)

data_long <- data %>%
  pivot_longer(cols = -Schooling_type, names_to = "Subject", values_to = "Mean_Score")

custom_colors <- c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd")

fill_order <- c("ACADEMIC", "TECHNICAL-PROFESSIONAL", "GENERAL EDUCATION", "ARTS", "ADULTS SCHOOL")


customize_labels_schooling <- function(value) {
  value <- as.character(value)
  custom_labels <- c("ACADEMIC" = "Academic",
                     "ADULTS SCHOOL" = "Adults School",
                     "ARTS" = "Arts",
                     "GENERAL EDUCATION" = "General Education",
                     "TECHNICAL-PROFESSIONAL" = "Technical-Professional")
  return(custom_labels[value])
}

customize_labels_subject <- function(value) {
  value <- as.character(value)
  custom_labels <- c("Spanish" = "Spanish",
                     "Mathematics" = "Math",
                     "Social_Sciences" = "Social Sciences",
                     "Natural_Sciences" = "Natural Sciences")
  return(custom_labels[value])
}

p <- ggplot(data_long, aes(x = reorder_within(Schooling_type, Mean_Score, Subject), y = Mean_Score, fill = Schooling_type, label = sprintf("%.2f", Mean_Score))) +
  geom_bar(stat = "identity") +
  geom_text(size = 2, position = position_stack(vjust = 0.5), color = "white") +  # Add Mean_Score as text
  labs(
    x = "Schooling Type's Mean Score",
    y = "Mean Score per Subject",
    fill = "School Type",
    title = "Performance Differences per School Type (2016-2020)",
    caption = "Source: National Examinations, Ministry of Education Dominican Republic"
  ) +
  theme(legend.title = element_text(),
        axis.text.y = element_blank()) +
  scale_fill_manual(values = custom_colors, breaks = fill_order) +
  facet_wrap(~Subject, ncol = 2, scales = "free", labeller = labeller(Schooling_type = customize_labels_schooling, Subject = customize_labels_subject)) +
  coord_flip()
p


```

In the context of national examinations, students' results are a combination of their school-based performance (weighted at 70%) and their national exam scores (weighted at 30%). Analyzing mean results by schooling type reveals expected variations within the educational system. However, identifying these trends is crucial for understanding which schooling types perform significantly better or worse than others. It is particularly important to highlight abnormal variations, as they open the door to explore the factors influencing academic outcomes.

**Academic Programs:** Students in Academic programs consistently achieve the highest mean scores, excelling in all subjects. This suggests a good academic foundation, contributing significantly to their overall national exam success.

**Adults School Programs:** Adults School programs show lower mean scores across subjects, indicating potential challenges or distinct academic expectations for students in these programs, potentially impacting their national exam performance.

**Arts Programs:** Arts programs perform moderately well, especially in Spanish. While not as high as Academic or General Education programs.

**General Education Programs:** General Education programs maintain well balanced mean scores across subjects, reflecting good academic performance.

**Technical-Professional Programs:** Technical-Professional programs exhibit high mean scores, particularly in Spanish. This signifies good academic performance.

These observations underscore the potential influence of educational programs on students' academic achievements and highlight variations that can help in understanding why certain schooling types perform better or worse on average. Identifying these trends is crucial for improving educational outcomes and addressing abnormal variations within the system.

### Correlation of Exam Performance and Location

```{r, echo = FALSE}
library(dplyr)

observed_regions <- c("01", "02", "09", "10", "15", "08")

region_subset <- HS_DR_Education %>%
  filter(`Regional` %in% observed_regions)

library(dplyr)

rural <- c("01", "02", "09")

region_subset <- region_subset %>%
  mutate(Urban_Rural = ifelse(`Regional` %in% rural, 0, 1))

```

```{r, echo = FALSE}

urban_schools <- region_subset %>% filter(Urban_Rural == 1)
rural_schools <- region_subset %>% filter(Urban_Rural == 0)

t_test_spanish <- t.test(urban_schools$Español, rural_schools$Español)
t_test_math <- t.test(urban_schools$Matemáticas, rural_schools$Matemáticas)
t_test_social <- t.test(urban_schools$Sociales, rural_schools$Sociales)
t_test_natural <- t.test(urban_schools$Naturales, rural_schools$Naturales)

```

```{r, echo = FALSE}
library(knitr)

t_test_results <- data.frame(
  Subject = c("Spanish", "Math", "Social Sciences", "Natural Sciences"),
  t = c(5.1032, 3.3673, 3.5812, 4.0391),
  df = c(4165.9, 4084.5, 4053.4, 4105.2),
  p_value = c(3.489e-07, 0.0007661, 0.000346, 5.464e-05),
  mean_diff = c(15.14413 - 14.42862, 14.60001 - 14.15433, 14.78045 - 14.25706, 14.74569 - 14.19774),
  lower_ci = c(0.4406269, 0.1861905, 0.2368591, 0.2819803),
  upper_ci = c(0.9903943, 0.7051646, 0.8099168, 0.8139205)
)

kable(t_test_results, caption = "T-Test: Exam Performance in Urban versus Rural Schools")
```

It should be noted that these results are not nationwide, but rather a comparison between selected regions, 3 urban and 3 rural. This subset has a total of 17,969 observations. 

**Spanish:**
The Two Sample t-test comparing the mean scores in Spanish between urban and rural schools yielded a statistically significant result (t = 5.1032, p-value = 0.0000003). The positive difference in means is approximately 0.716, and the 95 percent confidence interval for this difference ranges from 0.441 to 0.990. This suggests that, on average, students in urban schools have higher mean scores in Spanish compared to students in rural schools.

**Math:**
The Two Sample t-test comparing the mean scores in Mathematics between urban and rural schools also yielded a statistically significant result (t = 3.3673, p-value = 0.0007661). The positive difference in means is approximately 0.446, and the 95 percent confidence interval for this difference ranges from 0.186 to 0.705. This indicates that, on average, students in urban schools have higher mean scores in Mathematics compared to students in rural schools.

**Social Sciences:**
For Social Sciences, the Two Sample t-test produced a statistically significant result (t = 3.5812, p-value = 0.000346). The positive difference in means is approximately 0.523, and the 95 percent confidence interval for this difference ranges from 0.237 to 0.810. This implies that, on average, students in urban schools have higher mean scores in Social Sciences compared to students in rural schools.

**Natural Sciences:**
The Two Sample t-test comparing the mean scores in Natural Sciences also yielded a statistically significant result (t = 4.0391, p-value = 0.0000546). The positive difference in means is approximately 0.548, and the 95 percent confidence interval for this difference ranges from 0.282 to 0.814. This suggests that, on average, students in urban schools have higher mean scores in Natural Sciences compared to students in rural schools.

In summary, for each subject, the statistical analyses show that urban schools have higher mean scores compared to rural schools, and these differences are statistically significant. This may indicate variations in educational resources, teaching methods, or other factors between urban and rural schools that contribute to differences in student performance in these subjects. These findings are important for understanding educational disparities and may inform future research.

## Conclusion

The analysis of national examination results in the Dominican Republic highlights significant variations in mean scores across different schooling types and between urban and rural regions. Nationwide, Academic programs consistently achieve the highest performance, while Adults School programs exhibit lower mean scores. This is true across different academic subjects. The findings support the hypothesis that specific schooling types can influence student outcomes. Similarly, urban schools consistently outperform rural schools in all subjects, indicating potential disparities in educational resources and teaching methods. It is essential to acknowledge that even when the observed differences in mean scores between urban and rural schools may not be radical, they are still noteworthy. These variations, even if not extreme, raise important questions about the factors contributing to educational disparities.

There are several limitations and potential threats to inference in this analysis. Firstly, the scope of the second analysis (correlation) is limited to a subset of regions, which may not be representative of the entire Dominican Republic. This restricts the generalizability of the findings to the entire population. Additionally, the analysis relies on school's mean scores as the primary outcome measure, which may not capture the full complexity of student performance. 

Confounding variables, such as socioeconomic status or teacher quality, were not part of the dataset, but could significantly impact student performance. Future research could consider controlling for these potential confounders to provide a more accurate assessment of the relationships between schooling type, region, and academic outcomes.

If more time and resources were available, the analysis could be improved by conducting a nationwide study to ensure greater representativeness. Additionally, including a more comprehensive set of variables, such as socioeconomic factors and teacher characteristics, would enhance the understanding of the factors influencing student performance. 
